Accelerating Metagenomic Identification of DNA Sequences Using Artificial Neural Networks

### 2025-05-20 article/main.tex
article: change pipeline image
poprawki, RN

    In conclusion, the study highlights the potential of ANN-based methods to bridge the gap between precision and scalability in genetic taxonomy. Future research directions include enhancing the network’s flexibility for varying sequence lengths and exploring its application for dimensionality reduction in broader bioinformatics tasks.}

@@

    The rest of the paper is organized as follows. Section 2 defines the problem, presents the proposed approach, and provides detailed implementation details.
    Section 3 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis.
    Finally, section 4 concludes the paper by summarizing the main findings, discuss on the limtiations of proposed approach and suggesting potential directions for future research.

@@

        The core component of the proposed solution is a processing pipeline designed with flexibility in mind, allowing the integration of various DNA sequence clustering methods. The pipeline consists of a series of steps executed sequentially to perform metagenomic identification. A schematic representation of the pipeline is show in Figure~\ref{Picture:Pipeline}, where input and output data are represented by circles, and each processing step is represented by a rectangle. The pipeline is composed of the following five steps:

@@

        \begin{figure}[!htb]
            \begin{center}
                \includegraphics[width=0.8\textwidth]{picture_pipeline.png}
            \end{center}
            \caption{
                Diagram of the pipeline.
            }\label{Picture:Pipeline}
        \end{figure}

@@

    \subsection{\todo{czy to nie jest wybor reprezentanta?} Proposed Approach}

@@

        \subsection{Implementation Details}
            The solution consists of two components. The first is a library that contains components enabling the creation of pipelines for genetic material analysis in the form of metagenomic identification. The second component is a console application, which uses the library's components to allow for the creation and configuration of processing pipelines. Additionally, a web application was created, which allows users to submit analysis tasks through a web browser.

@@

                This sections presents the results of the experiments conducted to evaluate the proposed solution, which aims to improve performance while maintaining the quality of metagenomic identifacation. The study involved comparing the new approach with two classical methods using the same dataset. The evaluation focused on execution time and identification quality to assess the effectiveness of the proposed approach.

@@

        \subsection{Experiments}
            \subsubsection{Experiment 1: Metagenomic identification execution time}
                \paragraph{Objective} 
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

            \subsubsection{Experiment 2: Metagenomic identification quality}
                \paragraph{Objective}
                Examining the quality of metagenomic identification using the implemented methods in comparison to the metagenomic identification of all sequences.

@@

        \subsection{Performance Measures}
            Quality of metagenomic identification was measured using a modified Jaccard index, expressed by equation~\ref{Equation:Quality}. The measure was used to compare the quality of metagenomic identification performed using the implemented methods against metagenomic identification without the use of a processing pipeline.

@@

    \section{Conclusions}
        \subsection{Summary of Findings}
            The experiments demonstrated that the method based on artificial neural networks outperforms traditional approaches in terms of quality, while maintaining execution times comparable to those of the classical $k$-mer method. This highlights the potential of neural networks for improving performance without significant computational overhead.

        \subsection{Future Research Directions}
        Future research could focus on utilizing artificial neural networks as an independent tool for metagenomic identification. Another promising direction is the application of artificial neural networks for dimensionality reduction of genetic sequences, which may improve efficiency and performance in various analyses.

>>>>

    In conclusion, the study highlights the potential of ANN-based methods to bridge the gap between precision and scalability in genetic taxonomy.
    \todo{ We developed a tool with command-line interface, called Exquisitor that implements presented method.}

    Future research directions include enhancing the network’s flexibility for varying sequence lengths and exploring its application for dimensionality reduction in broader bioinformatics tasks.}

@@

    The rest of the paper is organized as follows. Section 2 defines the problem, presents the proposed approach, and provides detailed implementation details \todo{of new tool called Exquisitor}.
    Section 3 describes the experimental setup, including used datasets, the performance metrics for evaluation, and the results obtained from experiments with analysis.
    Finally, section 4 concludes the paper by summarizing the main findings, discuss on the limtiations of proposed approach and suggesting potential directions for future research.

@@

        The core component of \todo{the Exquisitor (proposed solution)} is a processing pipeline designed with flexibility in mind, allowing the integration of various DNA sequence clustering methods. The pipeline consists of a series of steps executed sequentially to perform metagenomic identification. A schematic representation of the pipeline is show in Figure~\ref{Picture:Pipeline}, where input and output data are represented by circles, and each processing step is represented by a rectangle. The pipeline is composed of the following five steps:

@@

        \begin{figure}[!htb]
            \begin{center}
                \includegraphics[width=1.0\textwidth]{picture_pipeline.png}
            \end{center}
            \caption{
                Diagram of the pipeline.
            }\label{Picture:Pipeline}
        \end{figure}

@@

    \subsection{\todo{Exquisitor method description}}

@@

        \subsection{Exquisitor implementation details}
            The solution consists of two components. The first is a library that contains components enabling the creation of pipelines for genetic material analysis in the form of metagenomic identification. The second component is a console application, which uses the library's components to allow for the creation and configuration of processing pipelines. Additionally, a web application was created, which allows users to submit analysis tasks through a web browser.

@@

                This sections presents the results of the experiments conducted to evaluate the proposed solution (Exquisitor), which aims to improve performance while maintaining the quality of metagenomic identifacation. The study involved comparing the new approach with two classical methods using the same dataset. The evaluation focused on execution time and identification quality to assess the effectiveness of the proposed approach.

@@

        \subsection{Experiments}
            \subsubsection{Metagenomic identification execution time \todo{using Exquisitor}}
                \paragraph{Objective}
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

            \subsubsection{Metagenomic identification quality}
                \paragraph{Objective}
                Examining the quality of metagenomic identification using the implemented methods in comparison to the metagenomic identification of all sequences.

@@

        \subsection{Exquisitor Performance Measures}
            Quality of metagenomic identification was measured using a modified Jaccard index, expressed by equation~\ref{Equation:Quality}. The measure was used to compare the quality of metagenomic identification performed using the implemented methods against metagenomic identification without the use of a processing pipeline.

@@

    \section{Conclusions}
        \subsection{Summary of Findings}
            The experiments demonstrated that \todo{the proposed method} based on artificial neural networks outperforms traditional approaches in terms of quality, while maintaining execution times comparable to those of the classical $k$-mer method. This highlights the potential of neural networks for improving performance without significant computational overhead.

        \subsection{Future Research Directions}
        Future research could focus on utilizing artificial neural networks as an independent tool for metagenomic identification. Another promising direction is the application of artificial neural networks for dimensionality reduction of genetic sequences, which may improve efficiency and performance in various analyses. \todo{zdanie: Planowane prace wdrożeniowe są skierowane w dostarczeniu interefsejsów do programu Exquisitor, pozwalając na jego integrację z istniejącymi popularnymi potokami (pipeline)}
